Adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2024-01-09 ms.tex
Final changes to address referee comments from round 1 of review

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.99\linewidth]{figures/deltaPhiTest.pdf}
    \caption{The log of the maximized phase difference $\log||\delta \Phi^I||_{\hat{a}, \hat{\Omega}}$ as described in Sec.~\ref{sec:traj} for different observation periods $T_\mathrm{obs}$. We can see that for any observation period, all the way up to 25 years, our phase spline maintains subradian phase accuracy when compared to direct numerical integration of the orbital phase.}
    \label{fig:obsTime}
\end{figure*}

@@

Next we quantify the impact of these interpolation errors on the gravitational wave phases. Recall from \eqref{eqn:adiabaticWaveform} that the phase of each gravitational wave mode $h_{\ell m}$ is proportional to the orbital phase $\Phi$. Therefore, the accuracy of the gravitational wave phases is set by the accuracy of $\Phi$. To quantify this accuracy, we first calculate the orbital phase accumulated over an inspiral with initial orbital frequency $\hat{\Omega}_0$ using our interpolated data,
\begin{align}
    \Delta \Phi^I &= \frac{1}{\epsilon}\left[\check{\Phi}^I\left(\check{t}_0 + \frac{\epsilon T}{M}; \hat{a} \right) - \check{\Phi}^I\left(\check{t}_0; \hat{a}\right) \right],
\end{align}
where $\check{t}_0 = \check{t}(\hat{\Omega}_0; \hat{a})$ and $T$ is the duration of the observed inspiral. Note that $\check{\Phi}^I$ is reconstructed from $\check{\Phi}^I_{N(2)}$ via \eqref{eqn:phaseNorm2}. We then compare $\Delta \Phi^I$ to the accumulated orbital phase $\Phi^\mathrm{ODE}$ obtained by directly integrating \eqref{eqn:eom1} with the initial conditions $\hat{\Omega}(\check{t}=0) = \hat{\Omega}_0$ and $\check{\Phi}(\check{t}=0) = 0$. The interpolation error is then estimated by the difference $\delta \Phi^I = |\Delta\Phi^I - \Phi^\mathrm{ODE}|$. Next, we find the maximum value of $\delta \Phi^I$ for a range of $\hat{a}$ values and initial orbital frequencies $\hat{\Omega}_0$. We denote this maximized difference as $||\delta \Phi^I||_{\hat{a},\hat{\Omega}}$. In Figure \ref{fig:obsTime} we plot $||\delta \Phi^I||_{\hat{a},\hat{\Omega}}$ as a function of mass-ratio $10^{-7} \leq \epsilon \leq 10^{-2}$ and massive black hole mass $10^{4} \leq M \leq 10^{8}$ for binaries observed for $T = [0.5, 2, 5, 10, 25]$ years. We exclude binaries with secondary masses $\mu < M_\odot$. We measure maximum phase errors of $0.0097$, $0.0186$, $0.0307$, $0.0430$, and $0.0642$ for the observation periods $0.5$, $2$, $5$, $10$, and $25$ years, respectively. These phase errors are reduced by about a factor of two if we only focus on inspirals with initial frequencies $\hat{\Omega} \geq 0.013$ (corresponding to $r_0 \lesssim 18 M)$ and binaries with smaller bodies of mass $\mu \geq 2 M_\odot$. Because the $(\ell, m) = (2,2)$ mode dominates each gravitational wave signal, $2 \Delta\Phi^I$ provides a reasonable estimate of the gravitational wave phase error. Consequently, our interpolation errors most likely meet the subradian phase accuracy requirements for realistic space-based mHz gravitational wave observations, which we further verify in Sec.~\ref{sec:errors}.

@@

To demonstrate the accuracy of our model, we compare our waveforms with those produced by FEW for $\hat{a} = e_0 = 0$. As a visual demonstration, in Figure \ref{fig:compareFEW} we plot $h_+$ for an EMRI with source properties $(M, \mu, p_0) = (10^6 M_\odot, 10 M_\odot, 12.05)$ observed in the SSB frame for four years. The strain calculated by FEW is given by the dashed lines, while the \texttt{bhpwave} result is given by the solid lines. The waveforms are aligned to agree at time $t=0$ (the beginning of the observed waveform). We zoom in on the waveform at three time windows placed near the beginning, middle, and end of the waveform. Across all three periods there is minimal disagreement between the two models. In the bottom panel of Figure \ref{fig:compareFEW}, we also plot the phase difference between the two models as a function of time. The two waveforms maintain subradian agreement over all five years of the signal.

@@

As a final test, we inject artificial noise into our original flux data and then recompute all of the trajectory splines to create a new biased waveform model, which we denote by $h_E$. The flux is modified via the correction
\begin{align}
    \mathcal{F}_E \rightarrow\mathcal{F}^\delta_E = \frac{1 + (1 + \delta)\frac{1247}{336}\hat{\Omega}^{2/3}}{1 + \frac{1247}{336}\hat{\Omega}^{2/3}} \mathcal{F}_E,
\end{align}
where we set $\delta = 10^{-5}$. The post-Newtonian expansion of $\mathcal{F}^\delta_E$ is
\begin{align}
    \mathcal{F}^\delta_E \sim \frac{32}{5}\hat{\Omega}^{10/3}\left[1 - \frac{1247}{336}(1 - \delta) \hat{\Omega}^{2/3} + O(\hat{\Omega}) \right].
\end{align}
Setting $\delta = 0$, we recover the first two post-Newtonian orders of the energy flux (e.g., see (178) in \cite{SasaTago03}), while $\delta = 10^{-5}$ adds a slight perturbation at the first (subleading) post-Newtonian order. From a post-adiabatic standpoint, this error is on the same order as 1PA corrections for systems with mass ratios $\mu/M \sim \delta$.

@@

To confirm the numerical accuracy of our interpolated data, we perform a series of comparisons and self-consistency checks. First we assess the accuracy of our flux interpolant by downsampling the data. Let $f^I_{(n,m)}(x,y)$ denote a bicubic spline that interpolates the function $f(x,y)$ using $(2^n+1) \times (2^m+1)$ grid points in $(x,y)$. Thus, in this notation our fully-sampled interpolant $\mathcal{F}_N^I(\alpha, \chi)$ is denoted by $\mathcal{F}_{N(8,7)}^I(\alpha, \chi)$, because it is constructed from a $257\times 128$ grid of flux values. We then estimate the fractional interpolation error for the flux spline $\mathcal{F}_{N(N',M')}^I$ via
\begin{align}
    \Delta^\mathcal{F}_{(N',M')}(\alpha, \chi) = \left|1 - \frac{\mathcal{F}_{N(N',M')}^I(\alpha, \chi)}{\mathcal{F}_{N(N'+1,M'+1)}^I(\alpha, \chi)} \right|.
\end{align}
By downsampling the data, we calculate $\Delta^\mathcal{F}_{(7,6)}$, $\Delta^\mathcal{F}_{(6,5)}$, and $\Delta^\mathcal{F}_{(5,4)}$. We plot $\Delta^\mathcal{F}_{(7,6)}$ in Figure \ref{fig:fluxSplineError}, demonstrating that $\Delta^\mathcal{F}_{(7,6)}<10^{-6}$ across the entire domain. To estimate the interpolation error of our fully-sampled spline, $\Delta^\mathcal{F}_{(8,7)}$, we examine the convergence rate of $|| \Delta^\mathcal{F}_{(n+1,n)} ||_\infty$ in Figure \ref{fig:fluxSplineConvergence}. We find that $|| \Delta^\mathcal{F}_{(n+1,n)} ||_\infty \propto 2^{-4n}$, which is consistent with the standard fourth-order scaling for cubic spline errors (error $ \sim \Delta x^4$ for grid spacing $\Delta x$). Provided this scaling holds true as we increase the sampling rate, we estimate that $\Delta^\mathcal{F}_{(8,7)} \lesssim 5\times 10^{-8}$.

>>>>

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.99\linewidth]{figures/deltaPhiTest.pdf}
    \caption{The logarithm of the maximized phase difference $\log||\delta \Phi^I||_{\hat{a}, \hat{\Omega}}$ as described in Sec.~\ref{sec:traj} for different observation periods $T_\mathrm{obs}$. We can see that for any observation period, all the way up to 25 years, our phase spline maintains subradian phase accuracy when compared to direct numerical integration of the orbital phase.}
    \label{fig:obsTime}
\end{figure*}

@@

Next we quantify the impact of these interpolation errors on the gravitational wave phases. Recall from \eqref{eqn:adiabaticWaveform} that the phase of each gravitational wave mode $h_{\ell m}$ is proportional to the orbital phase $\Phi$. Therefore, the accuracy of the gravitational wave phases is set by the accuracy of $\Phi$. To quantify this accuracy, we first calculate the orbital phase accumulated over an inspiral with initial orbital frequency $\hat{\Omega}_0$ using our interpolated data,
\begin{align}
    \Delta \Phi^I &= \frac{1}{\epsilon}\left[\check{\Phi}^I\left(\check{t}_0 + \frac{\epsilon T}{M}; \hat{a} \right) - \check{\Phi}^I\left(\check{t}_0; \hat{a}\right) \right],
\end{align}
where $\check{t}_0 = \check{t}(\hat{\Omega}_0; \hat{a})$ and $T$ is the duration of the observed inspiral. Note that $\check{\Phi}^I$ is reconstructed from $\check{\Phi}^I_{N(2)}$ via \eqref{eqn:phaseNorm2}. We then compare $\Delta \Phi^I$ to the accumulated orbital phase $\Phi^\mathrm{ODE}$ obtained by directly integrating \eqref{eqn:eom1} with the initial conditions $\hat{\Omega}(\check{t}=0) = \hat{\Omega}_0$ and $\check{\Phi}(\check{t}=0) = 0$. The interpolation error is then estimated by the difference $\delta \Phi^I = |\Delta\Phi^I - \Phi^\mathrm{ODE}|$. Next, we find the maximum value of $\delta \Phi^I$ for a range of $\hat{a}$ values and initial orbital frequencies $\hat{\Omega}_0$. We denote this maximized difference as $||\delta \Phi^I||_{\hat{a},\hat{\Omega}}$. In Figure \ref{fig:obsTime} we plot $||\delta \Phi^I||_{\hat{a},\hat{\Omega}}$ as a function of mass-ratio $10^{-7} \leq \epsilon \leq 10^{-2}$ and massive black hole mass $10^{4} \leq M \leq 10^{8}$ for binaries observed for $T = [0.5, 2, 5, 10, 25]$ years. We exclude binaries with secondary masses $\mu < M_\odot$. We measure maximum phase errors of $0.0097$, $0.0186$, $0.0307$, $0.0430$, and $0.0642$ for the observation periods $0.5$, $2$, $5$, $10$, and $25$ years, respectively. These phase errors are reduced by about a factor of two if we only focus on inspirals with initial frequencies $\hat{\Omega} \geq 0.013$ (corresponding to $r_0 \lesssim 18 M)$ and binaries with smaller bodies of mass $\mu \geq 2 M_\odot$. Because the $(\ell, m) = (2,2)$ mode dominates each gravitational wave signal, $2 \Delta\Phi^I$ provides a reasonable estimate of the gravitational wave phase error. Based on this approximation, our interpolation errors most likely meet the subradian gravitational wave phase accuracy requirements for realistic space-based mHz gravitational wave observations, which we further verify in Sec.~\ref{sec:errors}.

@@

To demonstrate the accuracy of our model, we compare our waveforms with those produced by FEW for $\hat{a} = e_0 = 0$. As a visual demonstration, in Figure \ref{fig:compareFEW} we plot $h_+$ for an EMRI with source properties $(M, \mu, p_0) = (10^6 M_\odot, 10 M_\odot, 12.05)$ observed in the SSB frame for four years. The strain calculated by FEW is given by the dashed lines, while the \texttt{bhpwave} result is given by the solid lines. The waveforms are aligned to agree at time $t=0$. We zoom in on the waveform at three time windows placed near the beginning, middle, and end of the waveform. Across all three periods there is minimal disagreement between the two models. In the bottom panel of Figure \ref{fig:compareFEW}, we also plot the phase difference between the two models as a function of time. The two waveforms maintain subradian agreement over all five years of the signal.

@@

As a final test, we inject artificial noise into our original flux data and then recompute all of the trajectory splines to create a new biased waveform model, which we denote by $h_E$. The flux is modified via the replacement
\begin{align}
    \mathcal{F}_E \rightarrow\mathcal{F}^\delta_E = \frac{1 + (1 + \delta)\frac{1247}{336}\hat{\Omega}^{2/3}}{1 + \frac{1247}{336}\hat{\Omega}^{2/3}} \mathcal{F}_E,
\end{align}
where we set $\delta = 10^{-5}$. With this rescaling, the post-Newtonian expansion of $\mathcal{F}^\delta_E$ is
\begin{align}
    \mathcal{F}^\delta_E \sim \frac{32}{5}\hat{\Omega}^{10/3}\left[1 - \frac{1247}{336}(1 - \delta) \hat{\Omega}^{2/3} + O(\hat{\Omega}) \right].
\end{align}
By setting $\delta = 0$ we recover the first two post-Newtonian orders of the energy flux (e.g., see (178) in \cite{SasaTago03}), while setting $\delta = 10^{-5}$ adds a slight perturbation at the first (subleading) post-Newtonian order.
From a post-adiabatic standpoint, this error is on the same order as 1PA corrections for systems with mass ratios $\mu/M \sim \delta$.

@@

To confirm the numerical accuracy of our interpolated data, we perform a series of comparisons and self-consistency checks. First we assess the accuracy of our flux interpolant by downsampling the data. Let $f^I_{(n,m)}(x,y)$ denote a bicubic spline that approximates a function $f(x,y)$ by interpolating sampled values of $f$ on a rectilinear $(2^n+1) \times (2^m+1)$ grid in $(x,y)$. Thus, in this notation our fully-sampled interpolant $\mathcal{F}_N^I(\alpha, \chi)$ can also be expressed as $\mathcal{F}_{N(8,7)}^I(\alpha, \chi)$, because it is constructed from a $257\times 128$ grid of flux values in $(\alpha, \chi)$. We then estimate the fractional interpolation error for the flux spline $\mathcal{F}_{N(N',M')}^I$ via
\begin{align}
    \Delta^\mathcal{F}_{(N',M')}(\alpha, \chi) = \left|1 - \frac{\mathcal{F}_{N(N',M')}^I(\alpha, \chi)}{\mathcal{F}_{N(N'+1,M'+1)}^I(\alpha, \chi)} \right|.
\end{align}
By downsampling the data, we calculate $\Delta^\mathcal{F}_{(7,6)}$, $\Delta^\mathcal{F}_{(6,5)}$, and $\Delta^\mathcal{F}_{(5,4)}$. We plot $\Delta^\mathcal{F}_{(7,6)}$ in Figure \ref{fig:fluxSplineError}, demonstrating that $\Delta^\mathcal{F}_{(7,6)}<10^{-6}$ across the entire domain. To estimate the interpolation error of our fully-sampled spline, $\Delta^\mathcal{F}_{(8,7)}$, we examine the convergence rate of $|| \Delta^\mathcal{F}_{(n+1,n)} ||_\infty$ in Figure \ref{fig:fluxSplineConvergence}. We find that $|| \Delta^\mathcal{F}_{(n+1,n)} ||_\infty \propto 2^{-4n}$, which is consistent with the standard fourth-order scaling for cubic spline errors (error $ \sim \Delta x^4$ for grid spacing $\Delta x$). Provided this scaling holds true as we increase the sampling rate, we estimate that $\Delta^\mathcal{F}_{(8,7)} \lesssim 5\times 10^{-8}$.
